Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-04 notebooks/expected_and_actual_counts_ratio.Rmd
Jitter plot to verify analyses and fix Bonferroni correction

---
title: "Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades"
output: html_document
date: "`r format(Sys.Date(), '%d.%m.%Y')`"
---

@@

[INSERT] Discuss purpose of notebook and steps to achieve that goal.

First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using either access to the server or from github.com.

@@

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

For the analyses conducted in this notebook, subset the data by the country(countries) the sequence data came from. Additionally, remove low-quality mutations or branches with abnormally high mutation rates. Finally, create a column with mutation type (there are 12 possible mutations).

@@

To compare mutations between phylogenetic clades, this notebook will plot volcano plots with p-value on the Y-axis and fold-change on the X-axis. Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated. Mutations with a small count (i.e., 1) are going to be affected by the pseudocount more than mutations with a large count (i.e., 100).

@@

The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).

```{r round_expected}
# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
```

## Characterize the Filtered Data

@@

# How many phylogenetic clades?
length(unique(filtered_SARS_mutation$clade))

@@

## Visualize the Filtered Data

@@

@@

With such a large data set, it will take potentially hours to calculate the p-values for clade comparisons. To decrease the amount of time it takes to conduct analyses, create a subset of the data with only the clades to be compared.

# Change column names so that clade 1 and clade 2 data frames can be combined
colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"

# Merge the two data frames by mutation so that each row represents one comparison.
filtered_comparisons <- merge(clade1, clade2, by="nt_mutation")
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'

# Save and load variable because calculating p-values takes a long time.
'save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))'
load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
```

Combine the p-values calculated using the Fisher's Exact test with the data frame of filtered comparisons.

```{r combine_p_values}
# Combine p-value list with the clade comparison list.
comparison_p_values <- cbind(filtered_comparisons, p_values)
```

@@

```{r p_value_correction}
# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_value_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
```

@@

Use a function to calculate fold-change using the observed-expected ratios calculated with a pseudocount. This value will be the X-axis for the volcano plot. P-values clustering less than zero indicate a difference in degree of selection for one clade, whereas p-values clustering greater than zero indicate a similar difference in degree of selection but for the other clade.

```{r fold-change}
# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
# Double check calculation with Jesse's paper
comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
```

Plot the volcano plot with -log10 p-values on the Y-axis and log fold-change on the X-axis.

```{r volcano_plot}
# Only plot data for the two clades being compared.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
  geom_point()+
  theme_bw()
```

## Verify Analyses Using Analyses With Expected Outcomes

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)
verify_analyses <- comparison_p_values
verify_analyses <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
nb <- sqrt(nrow(verify_analyses))
bw <- (max(verify_analyses$p_value_corrected)-min(verify_analyses$p_value_corrected))/nb
ggplot(verify_analyses, aes(x=p_value_corrected))+
  geom_histogram(bins = nb, binwidth = bw)
```

```{r alternative, echo=FALSE}
# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds

# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
#filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")

# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
#clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
#rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
#rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))

# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
#comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))

# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
#filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]

# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
#filtered_comparisons$id <- c(1:nrow(filtered_comparisons))

# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
#clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
#clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")

# Change column names so that clade 1 and clade 2 data frames can be combined
#colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
#colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")

# Order clade 1 and clade 2 data frames by id and combine
#fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
#fisher_data[,which(colnames(fisher_data) == "id")] <- NULL

# Identify clades from the two columns in the comparison matrix and combine into one column.
#comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))

# Count the number of rows for each clade (i.e., number of hypothesis tests).
#clade.table <- table(comparison_p_values$clades)

# Add a column for the number of hypothesis tests conducted for each pair of clades.
#comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
```

>>>>

---
title: "Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades"
output: html_document
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

@@

The purpose of this notebook is to generate volcano plots comparing any two SARS-CoV-2 phylogenetic clades. Using the volcano plots, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified. Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing the 78 clade comparisons to be computed in parallel.

Listed below are the general steps the notebook uses to generate a volcano plot:\
1. Read in the SARS-CoV-2 data containing observed and expected mutation counts.\
2. Set parameters for how to compute comparisons, including what clades to compare and what countries to accept data from.\
3. Filter data based on parameters.\
4. Restructure data such that there are columns for clade 1 variables and clade 2 variables.\
5. Using contingency table analyses (e.g., Fisher's Exact test), calculate p-values for each mutation at every nucleotide site between the two clades.\
6. Adjust p-values to account for the large number of hypothesis tests.\
7. Using the ratio of observed counts to expected counts for each clade and a pseudocount, calculate log2 fold-change.\
8. Generate a volcano plot with log2 fold-change on the X-axis and -log10 p-values on the Y-axis.\
9. Verify the notebook is computing p-values correctly by generating plots with expected outcomes (e.g., synonymous mutations).\

First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using access to the server or from the Bloom Lab github.com.

@@

# Identify the two clades that will be included in comparisons.
clades <- c("21J","22B")

@@

For the analyses conducted in this notebook, subset the data by the country (countries) the sequence data came from. Additionally, remove low-quality mutations or branches with abnormally high mutation rates.

@@

To compare mutations between phylogenetic clades, this notebook will plot volcano plots with -log10 p-value on the Y-axis and log2 fold-change on the X-axis. Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated. Mutations with a small count (i.e., 1) are going to be affected by the pseudocount more than mutations with a large count (i.e., 100).

@@

Contingency table analyses can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).

```{r round_expected}
# Round expected counts because contingency table analyses (e.g., Fisher's Exact test) can only analyze integers.
filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
```

## Describe and Characterize the Data

@@

@@

## Visualize the Data

@@

## Restructure Data for Analyses

With such a large data set, it could take hours to calculate the p-values for all clade comparisons. To decrease the amount of time it takes to conduct analyses, this notebook will only compare two clades at a time and the script can be run in parallel to generate all comparisons at once. Therefore, filter the data to only include mutation data from the two clades being compared.

To quickly calculate the p-value for each contingency table analysis and ensure the user can easily understand what is being compared in each hypothesis test, restructure the data such that all of the variables necessary to run the statistical test are in one row. That includes observed and expected counts for clade 1 and observed and expected counts for clade 2.

# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by="nt_mutation")
```

@@

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["clade1_actual_count"]),as.numeric(c["clade1_rounded_expected"])),c(as.numeric(c["clade2_actual_count"]),as.numeric(c["clade2_rounded_expected"]))))$p.value}))

# Save and load variable because calculating p-values may take a long time.
#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
```

Combine the p-values calculated using the contingency table analyses with the data frame of comparisons.

```{r combine_p_values}
# Combine the p-value list with the comparison list.
comparison_p_values <- cbind(comparisons, p_values)
```

@@

```{r p_value_correction}
# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
comparison_p_values$p_value_corrected <- comparison_p_values$p_value*nrow(comparison_p_values)
```

@@

Calculate fold-change using the observed-expected ratios calculated with a pseudocount. Fold-change is calculated as log2 of the clade 1 ratio divided by the clade 2 ratio. Therefore, if fold-change is positive then the clade 1 ratio is greater than the clade 2 ratio, and vise versa if fold-change is negative.

```{r fold-change}
# Calculate fold change for plotting volcano plots.
comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
```

Generate a volcano plot with -log10 p-values on the Y-axis and log2 fold-change on the X-axis.

```{r volcano_plot, echo=FALSE}
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
  geom_point()+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()
```

## Verify Analyses

To verify the analyses are generating expected results, compare the p-values for synonymous mutations to non-synonymous mutations. Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures. Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-value close to 0), there should be insignificant differences between clades on synonymous mutations (i.e., p-values close to 1).

```{r verify_analyses}
# sanity check analyses (distribution of synonymous p-values)
verify_analyses <- comparison_p_values
verify_analyses$synonymous <- paste(comparison_p_values$clade1_synonymous, comparison_p_values$clade2_synonymous, sep=".")

ggplot(verify_analyses, aes(x=synonymous, y=-log10(p_value_corrected)))+
  geom_jitter(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)
```
